---
title: "Seattle Flights"
author: "Stephen Kaluzny"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Data

The original data are from the 2009 ASA Data Expo (http://stat-computing.org/dataexpo/2009/).
The data consists of flight arrival and departure details
for all commercial flights within the USA,
from October 1987 to October 2008.
Flight data for November 2008 through December 2016 were downloaded from
The Department of Transportation's (DOT)
Bureau of Transportation Statistics (BTS) site
(https://www.transtats.bts.gov/OT_Delay/OT_DelayCause1.asp).

Flights to and from Seattle (SeaTac airport, airport code SEA) were extracted for each year and stored in ```data/SEA.csv```.
To save reading time, the extracted to and from data were then stored in R objects.

```{r data, eval=FALSE}
sea <- readr::read_csv("data/SEA.csv", col_types=list())
indxDest <- which(sea[["Dest"]] == "SEA")
FlightsToSeattle <- sea[indxDest, ]
indxOrig <- which(sea[["Origin"]] == "SEA")
FlightsFromSeattle <- sea[indxOrig, ]
saveRDS(FlightsToSeattle, "FlightsToSeattle.rds")
saveRDS(FlightsFromSeattle, "FlightsFromSeattle.rds")
```

```{r readdata, echo=FALSE}
FlightsToSeattle <- readRDS("FlightsToSeattle.rds")
FlightsFromSeattle <- readRDS("FlightsFromSeattle.rds")
```

Only flights with nonmissing ```ArrTime``` or ```DepTime``` were analyzed
```{r dataclean}
suppressPackageStartupMessages(library("dplyr"))
FlightsToSeattle <- FlightsToSeattle %>%
  filter(!is.na(ArrTime))
FlightsFromSeattle <- FlightsFromSeattle %>%
  filter(!is.na(DepTime))
```
The size of the data:
```{r size}
dim(FlightsToSeattle)
dim(FlightsFromSeattle)
```

Convert departure and arrival times to hour plus fraction of an hour.

```{r frachour}
FlightsToSeattle <- FlightsToSeattle %>%
  mutate(fracArrHour = ArrTime %/% 100 + (ArrTime %% 100) / 60)
FlightsFromSeattle <- FlightsFromSeattle %>%
  mutate(fracDepHour = DepTime %/% 100 + (DepTime %% 100) / 60)
```

Histogram of departure time:
```{r histdeparture}
b <- seq(0, 24, by=1/6)
hist(FlightsFromSeattle$fracDepHour, breaks=b, xlab="Hour",
  main="Flight Departure Time at SeaTac in Ten Minute Intervals",
  sub="Flights in 1987 through 2016",
  axes=FALSE)
abline(v = seq_len(23), lty=2, col="red", lwd=0.5)
axis(side=1, at=seq(0, 24, by=2))
axis(side=2)
```

Histogram of arrival time:
```{r histarrival}
b <- seq(0, 24, by=1/6)
hist(FlightsToSeattle$fracArrHour, breaks=b, xlab="Hour",
  main="Flight Arrival Time at SeaTac in Ten Minute Intervals",
  sub="Flights in 1987 through 2016",
  axes=FALSE)
abline(v = seq_len(23), lty=2, col="red", lwd=0.5)
axis(side=1, at=seq(0, 24, by=2))
axis(side=2)
```
